---
title: "Incarceration Report"
output: html_document
---

```{r setup, include=FALSE}
source("analysis.R")
library("knitr")
options(scipen = 999) #disable scientific notation formatting
```

## Overview
This is a brief report regarding incarceration trends in the United States
between 1970 and 2018 using Vera Institute's incarceration data set
([source](https://github.com/vera-institute/incarceration-trends)). The
variables that my analysis primarily focuses on are different measures of 
total jail population (different races), along with the corresponding jail
population rates. I also take a look at jail capacity and total jail
populations. 

## Summary Information
Here, I just chose a brief list of values that I personally was interested 
in finding. Just to be clear, there isn't any particular pattern or order to 
the information presented here; just things I was curious about. 

The data reveals that the county and state with the largest Native American
jail population in 2016 is `r place_top_na_jail_pop_2016`, 
with a population of `r top_native_jail_pop`. The year with the most overall
prison admissions was `r year_most_prison_adm` at 
`r round(num_prison_adm_largest_year, 0)` admissions. The average number of 
Black prisoners per prison in 2016 was `r round(avg_black_prison_pop_2016, 0)`. 
Finally, the total change in U.S prison populations from 2010 to 2018 is about
`r round(prison_pop_change_2000_to_2010, 0)` prisoners.

## Jail Incarceration Rate per Year of Top 5 Louisiana Counties
```{r, echo = FALSE, message=FALSE}
trend_plot
```

To clarify, the "Top 5 Louisiana Counties" are the top 5 counties by average
jail population over the time-horizon of the dataset (1970-2018). I included
this chart because I have often heard of Louisiana as a hotspot for crime in the 
U.S, and wanted to know more about the specific counties. Here are some of the 
patterns that emerged. Four of the five trend lines showed growth since
1980 before plateauing off and declining at around 2010. Particularly, the 
Orleans Parish saw dramatic increase in jail incarceration rate since 1980,
before rapidly declining in 2000. A possible contributor to the rapid decline
includes criminal justice reforms outlined in this 
[article](https://tinyurl.com/y3352pd5). Finally, Ouachita Parish is 
the outlier, with continued increases in jail rate, and no marked decline.  

## Jail Capacity versus Total Jail Population (2018)
```{r, echo = FALSE, message=FALSE, warning=FALSE}
comparison_plot
```

Although the results of the comparison may be mostly as expected, the chart
confirms intuition with data. One characteristic that good data 
scientists have is being skeptical about intuition -- and that's why 
why I created the chart. I wanted to see how close the relationship between
jail capacity and corresponding jail population was. As you can see,
the correlation is very strong and clearly linear. Particularly, the data
indicates that most jails are used nearly to capacity. This finding conforms
to news reports that we often hear about overcrowding of jails in the U.S. 
Another somewhat expected outcome demonstrated by the chart is that the 
larger the jail capacity the more likely for it to be located in a a more
urban setting.

## Proportion of Blacks in Jail Populations by State (2018)
```{r, echo = FALSE, message=FALSE}
map_plot
```

I included this chart to demonstrate the disproportionate population of 
Black/African Americans in U.S jails. Particularly the chart reveals that 
many states in the South, such as Alabama, Georgia, Mississippi, and Louisiana 
report average jail populations of 60% Black or greater. On the other hand,
the proportion of black people in the general population of these states 
generally hover between 25 to 30 percent. Therefore, Blacks are 
over-represented in jail populations for these states by two times
or more.